The Weakly Nonlinear Magnetorotational Instability in a local geometry

### 2017-03-20 paper/thingap_ref_response/thingap_mri_paper_ref_response_apj.tex
minor word changes; more edits on the bus.
added Oishi ORCID.
Edited simulations section; added some to conclusions.

\author{Jeffrey S. Oishi}
\affiliation{Department of Physics and Astronomy, Bates College, Lewiston, ME}
\affiliation{Department of Astrophysics, American Museum of Natural History, New York, NY}
\nocollaboration

@@

\begin{figure}[[h!]]
  \centering
  \includegraphics[width=\columnwidth]{growth_vs_B0.pdf}
  \caption{\edit1{Growth rate $\gamma$ as a function of background magnetic field strength $B_0$ at $\reym = \reym_c$, $k_z = k_c$, $\Pm = 10^{-3}$. Around the critical value $B_0 = 1.$, \emph{strengthening} $B_0$ tunes the system into instability, while \emph{decreasing} it leads to stability. The inset highlights the fact that $\gamma$ is determined by the maximum real eigenvalue of the system, which switches from one mode family to another as discussed in the text.}}
  \label{fig:growth_rate_vs_B0}
\end{figure}

@@

\edit1{We perform nonlinear, axisymmetric simulations of our fiducial setup using Dedalus.}

@@

In this paper we construct a weakly nonlinear analysis of the MRI using multiple scales analysis, leading to a real Ginzburg-Landau equation for the nonlinear amplitude, confirming the previous results of \citet{Umurhan:2007hs}. We also confirm their results for the scaling of the analytic saturation amplitude with $\Pm$. We extend their results by constructing a detailed force and inductive balance for the saturated $u_y$ and $B_z$ components. In doing so, we find the saturated state is a complex balance in which reduction of shear and amplification and redistribution of $B_z$ combine to saturate the instability. We also perform direct numerical simulations of the amplitude equation. We demonstrate that complex patterns can organize the flow on long length scales $Z$, though the maximum magnitude of the amplitude $\alpha$ is well predicted by the steady state solution. In Paper II, we make use of the techniques developed here to extend the weakly nonlinear analysis of the MRI to a full cylindrical geometry appropriate for a Taylor-Couette experiment.

@@

>>>>

\author[0000-0001-8531-6570]{Jeffrey S. Oishi}
\affiliation{Department of Physics and Astronomy, Bates College, Lewiston, ME}
\affiliation{Department of Astrophysics, American Museum of Natural History, New York, NY}
\nocollaboration

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\columnwidth]{growth_vs_B0.pdf}
  \caption{\edit1{Growth rate $\gamma$ as a function of background magnetic field strength $B_0$ at $\reym = \reym_c$, $k_z = k_c$, $\Pm = 10^{-3}$. Around the critical value $B_0 = 1.$, \emph{strengthening} $B_0$ tunes the system into instability, while \emph{decreasing} it leads to stability. The inset highlights the fact that $\gamma$ is determined by the maximum real eigenvalue of the system, which switches from one mode family to another as discussed in the text.}}
  \label{fig:growth_rate_vs_B0}
\end{figure}

@@

\edit1{Here, we make a preliminary test of our weakly nonlinear theory by comparing it to direct numerical simulation. Using Dedalus, we  solved the full, nonlinear equations \ref{eqset1} - \ref{eqset4} with all parameters ($\reym$, $Q$) equal to their critical values except the background magnetic field, which we set to $B_0 = 1 + \epsilon^2$. We thus drive the system MRI unstable in the same way as in our theory. The computational requirements of low $\Pm$ simulations are quite intense in both time and space. Despite being virtually smooth, the solutions require a resolution of $192 \times 1536$ grid points at $\Pm = 10^{-2}$. Because the system has such a small growth rate, it takes hundreds of orbits for the system to reach saturation, as compared to the few orbits typical of high $\reym$ simulations \citep[e.g.][]{2007MNRAS.378.1471L}. As a result, we make our comparison at $\Pm = 10^{-2}$, which provides a good tradeoff between probing relatively low $\Pm$ while keeping the computational time for these exploratry simulations modest. }

\edit1{We initialize the runs with the linear eigenvectors of the MRI unstable mode (also computed by Dedalus; see section~\ref{sec:wnl}) multiplied by an initial amplitude $A_0 = 10^{-3}$. Doing so requires considerably less run time, as the MRI unstable mode starts growing immediately from $A_0$. By contrast, initializing random noise in $\psi$ with amplitude $A_0$ would give the unstable mode a much smaller amplitude. Nevertheless, we have confirmed that simulations with eigenvector initial conditions have similar evolutions to those with noise initial conditions once each enter linear growth. }

\textbf{SUSAN: SIMULATION RESULTS GO HERE}

@@

In this paper we construct a weakly nonlinear analysis of the MRI using multiple scales analysis, leading to a real Ginzburg-Landau equation for the nonlinear amplitude, confirming the previous results of \citet{Umurhan:2007hs}. We also confirm their results for the scaling of the analytic saturation amplitude with $\Pm$. We extend their results by constructing a detailed force and inductive balance for the saturated $u_y$ and $B_z$ components. In doing so, we find the saturated state is a complex balance in which reduction of shear and amplification and redistribution of $B_z$ combine to saturate the instability. \edit1{We also perform numerical simulations of the amplitude equation and a direct numerical simulation of the MRI system.} \edit1{Using the former, we demonstrate that complex patterns can organize the flow on long length scales $Z$, though the maximum magnitude of the amplitude $\alpha$ is well predicted by the steady state solution. The latter show that there is rough agreement for both total energy and average angular momentum transport between the weakly nonlinear theory and simulation for a representative case at $\Pm = 10^{-2}$. We defer a full comparison between theory and simulation to later work.} In Paper II, we make use of the techniques developed here to extend the weakly nonlinear analysis of the MRI to a full cylindrical geometry appropriate for a Taylor-Couette experiment.

@@

\bibitem[Lesur \& Longaretti(2007)]{2007MNRAS.378.1471L} Lesur, G., \&
  Longaretti, P.-Y.\ 2007, \mnras, 378, 1471
